Deciding Bitvector Termination with Program Synthesis

### 2014-07-05 papers/termination/kalashnikov.tex
Synthesis algorithm

\subsection{The Synthesis Algorithm}
As safety of a \newC program is efficiently 
%the quantifier free first-order fragment of the logic is efficiently 
decidable and a termination proof is expressible as a ground term of the \newC logic, we use
Counterexample Guided Inductive Synthesis (CEGIS)~\cite{lezama-thesis,sketch} to
check the validity of the synthesis formula. 
The core of the CEGIS algorithm is the refinement loop shown in Fig.~\ref{fig:abstract-refinement} and
detailed in Algorithm~\ref{fig:abstract-refinement-code}.  
The algorithm is divided into two
components: {\sc synth} (see Figure~\ref{fig:synth-dfd}) and {\sc verif}.% described next.

Given a (non-)termination specification (in the synthesis fragment)    
and a finite set of test inputs tracked so far, {\sc synth} generates a program \texttt{synth.c} asserting that some non-deterministic proof taken as input  
is incorrect on at least one of the tracked inputs.
Thus, \texttt{synth.c} is unsafe iff there is some (non-)termination proof that is correct for all of the tracked inputs.  
By using testing techniques, we search for such a proof, i.e. 
we synthesise a candidate (non-)termination proof that is correct for all the tracked inputs.  
The testing techniques we use are: model checking,
exhaustive search and genetic programming (see Section~\ref{sec:gp}).  
%For the latter 2 techniques, we combine SYNTH with another program
%(GP or EXPLICIT respectively) to create a program that searches for test vectors (proofs) that cause SYNTH
%to crash.
As \texttt{synth.c}
is a \newC program, we can check its safety by invoking a Bounded
Model Checker, such as {\sc cbmc} \cite{}.
We run the three available testing techniques %explicit-state model checker
in parallel and take the answer of whichever happens
to terminate first, stopping the other processes.

Subsequently, {\sc Verif} uses model checking or exhaustive search to verify the candidate proof by searching for an input on
which it does not satisfy the specification. If no such input
can be found, the candidate proof is correct.  Otherwise, the new input is added
to the set of tracked inputs and the CEGIS loop is reiterated.

@@

\coordinate (midspec) at ($(spec) !.5! (file)$);
\draw (midspec |- 0,-3) node [align=center] {Automatically generated specification \\ (written in C)};
%\draw (file |- 0,-3) node {Spec};
\draw (gcc |- 0,-3) node [align=center] {Test generator};
\draw (out |- 0,-3) node [align=center] {Proof \\ (written in $\mathcal{L}$)};

>>>>

\subsection{The Synthesis Algorithm}
As safety of a \newC program is efficiently 
%the quantifier free first-order fragment of the logic is efficiently 
decidable and a termination proof is expressible as a ground term of the \newC logic, we use
Counterexample Guided Inductive Synthesis (CEGIS)~\cite{lezama-thesis,sketch} to
check the validity of the synthesis formula. 
The core of the CEGIS algorithm is the refinement loop shown in Fig.~\ref{fig:abstract-refinement} and
detailed in Algorithm~\ref{fig:abstract-refinement-code}.  
The algorithm is divided into two
procedures: {\sc synth} (see Figure~\ref{fig:synth-dfd}) and {\sc verif}, which interact via
a finite set of test vectors {\sc inputs}, which is initialised to $\emptyset$.

The {\sc synth} procedure tries to find existential witnesses $(P, x_0)$ which satisfy the partial specification:
\[
 \exists P, x_0 . \forall x \in \text{\sc inputs} . \sigma(P, x_0, x)
\]

If {\sc synth} succeeds in finding a witness $(P, x_0)$, this witness is a candidate solution to the full
synthesis formula.  We pass this candidate solution to {\sc verif} which determines whether the candidate
does satisfy the specification on all inputs by checking satisfiability of the verification formula:
\[
 \exists x . \lnot \sigma(P, x_0, x)
\]

If this formula is unsatisfiable, the candidate solution is in fact a solution to the synthesis formula
and so the algorithm terminates.  Otherwise, the witness $x$ is an input on which the candidate solution fails
to meet the specification.  This witness $x$ is added to the {\sc inputs} set and the loop iterates again.

Concretely, {\sc synth} is implemented as shown in Figure~\ref{fig:synth-dfd}.  We start by taking the termination
specification and generating a C program which takes as input a candidate $(P, x_0)$ and asserts that the candidate
fails to meet the specification on at least on of the elements of {\sc inputs}.  So this program is unsafe iff there
is some candidate $(P, x_0)$ satisfying the specification for all the elements of {\sc inputs}.  Finding a new
candidate solution is therefore reduced to model checking this C program, which by construction is loop free.
The model checking is done using a combination of bounded model checking, explicit state enumeration and
genetic programming (GP).  The latter two techniques involve combining the previously generated C program,
along with code that searches for candidate solutions.  As well as efficiency, this has the side effect of
guaranteeing that we are verifying the exact semantics of the program under analysis as understood by a
compiler, rather than some ad-hoc abstraction of its semantics.  Finally, if any of the model checkers
finds a candidate solution it returns it.  The form of this candidate solution is yet another program,
this time written in the proof language $\mathcal{L}$.

Similarly, the {\sc verif} procedure is implemented by creating a C program that is unsafe iff
some input exists on which the candidate solution fails to meet the specification.  Again, symbolic
and explicit state model checking are used to decide the satisfiability of the verification formula.

@@

\coordinate (midspec) at ($(spec) !.5! (file)$);
\draw (midspec |- 0,-3) node [align=center] {Automatically generated specification \\ (written in C)};
%\draw (file |- 0,-3) node {Spec};
\draw (gcc |- 0,-3) node [align=center] {Model checker};
\draw (out |- 0,-3) node [align=center] {Proof \\ (written in $\mathcal{L}$)};
